
\section{Performance}

%A published benchmark (independent of our lab) found that
%\textsc{infernal} and other CM based methods were the most sensitive
%and specific tools for structural RNA homology search among the
%several that were tested \citep{Freyhult07}.  Results on a separate
%benchmark that we use during \textsc{infernal} development have been
%consistent with that conclusion \citep{NawrockiEddy07}.

A published benchmark (independent of our lab) \citep{Freyhult07} and
our own internal benchmark used for \textsc{infernal} development
\citep{NawrockiEddy07} both find that \textsc{infernal} and other CM
based methods are the most sensitive and specific tools for structural
RNA homology search among the several that were tested. Figure~1 shows
updated results of our internal benchmark comparing \textsc{infernal}
1.0 to the previous version (0.72) that was benchmarked in
\citet{Freyhult07}, and also to family-pairwise-search with BLASTN
\citep{Altschul97,Grundy98b}.  The sensitivity and specificity of
\textsc{infernal} 1.0 have greatly improved relative to 0.72. There
have been three relevant improvements in the implementation: a biased
composition correction to the raw log-odds scores, the use of the full
Inside log-likelihood scores (summed over all alignments) in place of
CYK maximum likelihood alignment scores, and the introduction of
approximate E-value estimates for the scores.

%\begin{comment}
The benchmark dataset used in Figure~1 
%is an improved version of our internal Rfam-based benchmark
%\citep{NawrockiEddy07}. Briefly, this benchmark 
was constructed as follows. The sequences of the seed
alignments of 503 Rfam (release 7) families were single linkage
clustered by pairwise sequence identity, and separated into two
clusters such that no sequence in one cluster is more than 60\%
identical to any sequence in the other. The larger of the two clusters
was assigned as the query (preserving their original Rfam alignment
and structure annotation), and the sequences in the smaller cluster
were assigned as true positives in a test set. We required a minimum
of five sequences in the query alignment. 51 Rfam families met these
criteria, yielding 450 test sequences which were embedded at random
positions in a 10 Mb ``pseudogenome''.  Previously we generated the
pseudogenome sequence from a uniform residue frequency distribution \citep{NawrockiEddy07}.
%Because base composition biases
%in the target sequence database cause the most serious problems in
%separating significant CM hits from noise, 
Here, we generated a more realistic pseudogenome sequence using a
15-state fully connected hidden Markov model (HMM) trained by
Baum-Welch expectation maximization \citep{Durbin98} on genome
sequence data from a wide variety of species.  Each of the 51 query
alignments was used to build a CM and search the pseudogenome, a
single list of all hits for all families were collected and ranked,
and true and false hits were defined (as described in
\citet{NawrockiEddy07}), producing the ROC curves in Figure~1.
%\end{comment}

\textsc{infernal} searches require a large amount of compute time
(Table~1). To alleviate this, \textsc{infernal} 1.0 implements two
rounds of filtering.  When appropriate, the HMM filtering technique
described by \citet{WeinbergRuzzo06} is applied
first with filter thresholds configured by \emph{cmcalibrate}
(occasionally a model with little primary sequence conservation cannot
be usefully accelerated by a primary sequence based filter).  The
query-dependent banded (QDB) CYK search algorithm is used as a second
filter with relatively tight bands ($\beta$= $10^{-7}$)
\citep{NawrockiEddy07}.  Any sequence fragments that survive the filters are
searched a final time with the Inside algorithm (again using QDB, but with looser
bands ($\beta$= $10^{-15}$)).
%The benchmark in Figure~1 shows results with and without the filters.  
In our benchmark, the default filters accelerate similarity search by
about 30-fold overall, while sacrificing a small amount of sensitivity
(Figure~1). This makes version 1.0 substantially faster than
0.72. \textsc{BLAST} is still orders of magnitude faster, but
significantly less sensitive than \textsc{infernal}. Further
acceleration remains a major goal of \textsc{infernal} development.
%Table~1 shows specific examples of running times for
%similarity searches for six RNA families of various sizes with and
%without filters.
%The acceleration
%gained from filters varies widely. In general, the more primary
%sequence conservation in a family, the more effective an HMM filter
%will be, although this is not always the case. 

The computational cost of CM alignment with the \emph{cmalign} program
has been a limitation of previous versions of
\textsc{infernal}. Version 1.0 now uses a constrained dynamic
programming approach first developed by \citet{Brown00}
that uses sequence specific bands derived from a first-pass HMM
alignment. This technique offers a dramatic speedup relative to
unconstrained alignment, especially for large RNAs such as small and
large subunit (SSU and LSU) ribosomal RNAs, which can now be aligned
in roughly 1 and 3 seconds per sequence, respectively (Table 1), as
opposed to 12 minutes and 3 hours in previous versions.
We expect this to be particularly useful in applications where
many large RNA sequences need to be aligned. One of the main ribosomal
RNA databases, RDP, has recently adopted \textsc{infernal} in its
pipeline \citep{Cole09}.

