
\section{Usage} 

A CM is built from a Stockholm format \citep{infguide03} multiple
sequence alignment (or single RNA sequence) with consensus secondary
structure annotation marking which positions of the alignment are
single stranded and which are base paired. CMs assign position
specific scores for the four possible residues at single stranded
positions, the sixteen possible base pairs at paired positions, and
for
%as well as position specific scores for
insertions and deletions. These scores are log-odds scores derived
from the observed counts of residues, base pairs, insertions and
deletions in the input alignment, combined with prior information
derived from structural ribosomal RNA alignments. Construction and
parameterization of CMs have been described in more detail elsewhere
\citep{Eddy94,infguide03,Eddy02b,NawrockiEddy07}.

\textsc{infernal} is composed of several programs that are used in
combination to build models, search databases, and align putative
homologs, following four basic steps:

\begin{enumerate}
\item Build a CM from a structural alignment with \emph{cmbuild}.

\begin{comment}
\emph{cmbuild} takes as input a structural multiple
RNA alignment in Stockholm format \citep{infguide03} and creates a CM
file that is used by other \textsc{infernal} programs.
\end{comment}
\begin{comment}
\emph{cmbuild} creates a CM file for use with other \textsc{infernal}
programs from an input structural multiple alignment in Stockholm
format \cite{infguide03}.
\end{comment}
\item Calibrate a CM for similarity search with \emph{cmcalibrate}.

\item Search databases for putative homologs to a CM with \emph{cmsearch}.

\begin{comment}
Given a CM file and a target database as input, 
\emph{cmsearch} searches the target database
for high scoring hits to the model and outputs alignments
of each hit in a BLAST-like format augmented with structure
annotation.
\end{comment}

\item Align putative homologs to a CM with \emph{cmalign}.
% to create a structurally annotated multiple alignment in Stockholm format.

\begin{comment}
\emph{cmalign} takes a CM file and a file of putative homologs, and
aligns the full length sequences to the model, creating a 
structurally annotated multiple alignment in Stockholm format.
\end{comment}

\end{enumerate}

The calibration step is optional and computationally expensive (10
hours for a CM of a typical RNA family), but is required to obtain
E-values that estimate the statistical significance of each hit in a
database search. \emph{cmcalibrate} will also determine appropriate
HMM filter thresholds for accelerating searches without an appreciable
loss of sensitivity. Each model only needs to be calibrated once.

%Some steps are unnecessary for some applications. For example, a user
%that wants only to generate alignments of previously defined
%homologous sequences, such as small subunit ribosomal RNA (SSU rRNA)
%sequences, would skip the calibration and search steps.

%For similarity search applications, where the goal is to identify new
%examples of a family, it is reasonable to iterate these steps, adding
%newly found homologs to the alignment and repeating the search as the
%detected range of the family expands. Just as with primary sequence
%profiles, the ability of CMs to detect remote homologs tends to
%increase as the diversity of known sequences in the query alignment
%increases.
