\section{Introduction}

\software{Infernal} is a software package that allows you to make
consensus RNA secondary structure profiles, and use them to search
nucleic acid sequence databases for homologous RNAs, or to create new
structure-based multiple sequence alignments.

To make a profile, you need to have a multiple sequence alignment of
an RNA sequence family, and the alignment must be annotated with a
consensus RNA secondary structure. The program \prog{cmbuild} takes an
annotated multiple alignment as input, and outputs a profile.

You can then use that profile to search a sequence database for homologs,
using the program \prog{cmsearch}.

You can also use the profile to align a set of unaligned sequences to
the profile, producing a structural alignment, using the program
\prog{cmalign}. This allows you to build hand-curated representative
alignments of RNA sequence families, then use a profile to
automatically align any number of sequences to that profile.  This
seed alignment/full alignment strategy combines the strength of
stable, carefully human-curated alignments with the power of automated
updating of complete alignments as sequence databases grow. This is
the strategy used to maintain the \database{Rfam} database of RNA
multiple alignments and profiles.

\software{Infernal} is comparable to \software{hmmer}
(\htmladdnormallink{hmmer.wustl.edu}{http://hmmer.wustl.edu}).  The
\software{hmmer} software package builds profile hidden Markov models
(profile HMMs) of multiple sequence alignments. Profile HMMs capture
only primary sequence consensus features. \software{Infernal} models
are profile stochastic context-free grammars (profile SCFGs).  Profile
SCFGs include both sequence and RNA secondary structure consensus
information.

\emph{Currently \software{Infernal} is really just an algorithm
testbed. Output is rudimentary, and many desired features are
missing.} Most importantly, \software{Infernal} is very slow and
CPU-intensive. You will probably need a large number of CPUs in order
to use it for serious work. Planned algorithmic improvements should
make it more practical in the future. We are making it available as a
fully documented package now, only because \software{Infernal} has
been pressed prematurely into service as the basis for constructing
and maintaining the \database{Rfam} database of structurally annotated
RNA multiple alignments \cite{Griffiths-Jones03}.










